---
title: "H2171_GSEA_GO"
author: "Nivitha Bhaskar"
date: "2025-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Use the DESEq2 output files to perform the GSEA analysis

#Loading necessary packages
```{r libraries}
library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
library(AnnotationDbi)
library(ggplot2)
library(enrichplot)
library(GO.db)
library(GOplot)
library(enrichplot)
library(KEGGREST)
library(ggrepel)
library(gridExtra)
library(stringr)
library(tidyr)
```

# loading differential gene exp results
```{r}
# List of the file names
file_names <- c("H2171_corin_ic50_final.csv", "H2171_corin_ic75_final.csv", 
                "H2171_ent_ic50_final.csv", "H2171_ent_ic75_final.csv", 
                "H2171_sp_ic50_final.csv", "H2171_sp_ic75_final.csv")

# Read all the CSV files into a list of data frames
df_list <- lapply(file_names, read.csv)

# Optionally, assign names to each element in the list for easy reference
names(df_list) <- c("H2171_corin_ic50", "H2171_corin_ic75", "H2171_ent_ic50", "H2171_ent_ic75", "H2171_sp_ic50", "H2171_sp_ic75")
```

# GSEA function
Function converts ENSEMBL to ENTREIZ ids, then runs both GO and KEGG

```{r}
run_gsea_ensembl <- function(df) {
  # Clean ENSEMBL IDs just in case there are version suffixes
  ens_ids <- gsub("\\..*", "", as.character(df$X))

  # Map to ENTREZ IDs
  df$entrez <- mapIds(org.Hs.eg.db,
                      keys = ens_ids,
                      column = "ENTREZID",
                      keytype = "ENSEMBL",
                      multiVals = "first")

  # Extract gene list (e.g., log2FoldChange or stat)
  gene_list <- df[, 3]
  names(gene_list) <- as.character(df$entrez)

  # Clean gene list
  gene_list <- sort(gene_list, decreasing = TRUE)
  gene_list <- gene_list[!duplicated(names(gene_list))]
  gene_list <- gene_list[!is.na(names(gene_list))]

  # Run gseGO
  gse_go <- gseGO(geneList = gene_list,
                      OrgDb = org.Hs.eg.db,
                      ont = "BP",
                      minGSSize = 100,
                      maxGSSize = 500,
                      pvalueCutoff = 0.05,
                      verbose = FALSE)
   # Run gseKEGG
  gse_kegg <- gseKEGG(geneList = gene_list,
                      organism = "hsa",
                      minGSSize = 10,
                      maxGSSize = 500,
                      pvalueCutoff = 0.05,
                      verbose = FALSE)

  # Return both results as a list
  return(list(GO = gse_go, KEGG = gse_kegg))

}

# Apply to all datasets
gsea_results <- lapply(df_list, run_gsea_ensembl)
names(gsea_results) <- names(df_list)
```

# Visualizing GSEA as dotplots
will be saved as separate pdf


```{r}
pdf("GSEA_2171_RNAseq.pdf", width = 10, height = 8)

for (name in names(gsea_results)) {
  # Plot GO results
  p1<-(dotplot(gsea_results[[name]]$GO,
                showCategory = 20,
                title = paste("Top 20 Enriched GO Terms in", name)))
  print(p1)

  # Plot KEGG results
  p2<-(dotplot(gsea_results[[name]]$KEGG,
                showCategory = 20,
                title = paste("Top 20 Enriched KEGG Pathways in", name)))
  print(p2)
}

dev.off()
```

# Viewing results as Gsea plots
```{r}
# Define conditions
conditions <- c("H2171_corin_ic50", "H2171_corin_ic75", "H2171_ent_ic50", 
                "H2171_ent_ic75", "H2171_sp_ic50", "H2171_sp_ic75")

# Initialize list to collect plots
plot_list <- list()

# Loop through each condition
for (cond in conditions) {
  
  # Access the GSEA KEGG result
  gsea_res <- gsea_results[[cond]]$KEGG
  
  # Skip if KEGG results are missing
  if (is.null(gsea_res) || nrow(as.data.frame(gsea_res)) == 0) {
    cat("Skipping:", cond, "- no KEGG data\n")
    next
  }
  
  # Filter significant pathways (adjusted p < 0.05)
  filtered_result <- gsea_res@result %>%
    filter(p.adjust < 0.05)
  
  if (nrow(filtered_result) == 0) {
    cat("Skipping:", cond, "- no significant KEGG terms (padj < 0.05)\n")
    next
  }
  
  # Select top 10 significant pathways
  top_pathways <- filtered_result %>%
    arrange(p.adjust) %>%
    slice_head(n = 10) %>%
    pull(ID)
  
  # Plot the top 10 pathways
  p <- gseaplot2(gsea_res, 
                 geneSetID = top_pathways, 
                 title = paste("Top 10 Significant KEGG Pathways -", cond),
                 base_size = 12)
  
  # Add to list
  plot_list[[cond]] <- p
}

# Combine all plots into one PDF file
pdf("H2171_GSEA_KEGG_Significant_Top10.pdf", width = 17, height = 12)
for (p in plot_list) {
  print(p)
}
dev.off()

cat("✅ Significant KEGG GSEA plots saved to 'H2171_GSEA_KEGG_Significant_Top10.pdf'\n")
```

# for negatively significant
```{r}

# Define conditions
conditions <- c("H2171_corin_ic50", "H2171_corin_ic75", "H2171_ent_ic50", 
                "H2171_ent_ic75", "H2171_sp_ic50", "H2171_sp_ic75")

# Initialize list to store plots
plot_list <- list()

# Loop through each condition
for (cond in conditions) {
  
  gsea_res <- gsea_results[[cond]]$KEGG
  
  if (is.null(gsea_res) || nrow(gsea_res@result) == 0) {
    warning(paste("No valid KEGG result for:", cond, "- skipping."))
    next
  }

  # Filter for negatively enriched pathways
  neg_enriched <- gsea_res@result %>%
    filter(NES < -1, p.adjust < 0.05) %>%
    arrange(p.adjust) %>%
    slice_head(n = 10)
  
  if (nrow(neg_enriched) == 0) {
    warning(paste("No negatively enriched pathways in", cond))
    next
  }

  top_neg_pathways <- neg_enriched$ID

  # Generate plot
  p <- gseaplot2(gsea_res, 
                 geneSetID = top_neg_pathways, 
                 title = paste("Top Negatively Enriched KEGG Pathways -", cond),
                 base_size = 11) 
  
  # Store in list
  plot_list[[cond]] <- p
}

# Save all plots into a single PDF file
pdf("All_Neg_Enriched_KEGG_GSEA_Plots.pdf", width = 17, height = 12)
for (p in plot_list) {
  print(p)
}
dev.off()

```

# Digging deeper into the enriched genes
After looking at the gseaplots, found hematopoetic  biosynthesis pathway to be 
strongly negatively enriched, therefore looking at the specific genes

```{r}
# Get gene information for pathway hsa00900 
kegg_genes <- keggGet("hsa04640")[[1]]$GENE

# The list includes Entrez ID and gene description, alternating
gene_ids <- kegg_genes[seq(1, length(kegg_genes), 2)]  # Get only the Entrez IDs

# Convert Entrez IDs to gene symbols
gene_symbols <- bitr(gene_ids, fromType = "ENTREZID",
                     toType = "SYMBOL", OrgDb = org.Hs.eg.db)

# View the gene symbols
print(gene_symbols)


# Your list of gene symbols from the KEGG pathway
hema_genes <- gene_symbols$SYMBOL

# Subset DESeq2 results
# Subset each dataset for terpenoid pathway genes
hema_subsets <- lapply(df_list, function(df) {
  df[df$symbol %in% hema_genes, ]
})

# Custom volcano function using your styling
make_custom_volcano <- function(df, dataset_name) {
  # Add required columns
  df$log2fc <- df$log2FoldChange
  df$pval <- df$padj
  df$delabel <- df$symbol
  
  # Classify for coloring
  df$diffexpressed <- "Not significant"
  df$diffexpressed[df$log2fc > 1 & df$padj < 0.05] <- "Upregulated"
  df$diffexpressed[df$log2fc < -1 & df$padj < 0.05] <- "Downregulated"
  
  # Plot
  ggplot(data = df, aes(x = log2fc, y = -log10(pval), col = diffexpressed, label = delabel)) +
    geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
    geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
    geom_point(size = 2) +
    scale_color_manual(
      values = c("Downregulated" = "#00AFBB", "Not significant" = "grey", "Upregulated" = "#bb0c00"),
      labels = c("Downregulated", "Not significant", "Upregulated")
    ) +
    coord_cartesian(ylim = c(0, 25), xlim = c(-7, 7)) +
    labs(
      color = 'Expression',
      x = expression("log"[2]*"FC"),
      y = expression("-log"[10]*"adjusted p-value")
    ) +
    scale_x_continuous(breaks = seq(-10, 10, 2)) +
    ggtitle(paste("Hematopoietic cell lineage -", dataset_name)) +
    geom_text_repel(max.overlaps = Inf, size = 2)+
    theme_classic()
}

# Apply to each dataset in terpenoid_subsets
custom_hema_plots <- mapply(
  FUN = make_custom_volcano,
  df = hema_subsets,
  dataset_name = names(hema_subsets),
  SIMPLIFY = FALSE
)


# Or use gridExtra to show multiple
grid.arrange(grobs = custom_hema_plots[1:6], ncol = 2)  # change range/ncol as needed
```

# Visualizing circular plots

First extracting the go results
```{r}
# Define output PDF
pdf("H2171_GOChord_plots.pdf", width = 20, height = 20)

# Loop through each element of gsea_results
for (s in names(gsea_results)) {
  cat("Processing:", s, "\n")

  # STEP 1: Extract and clean GSEA result
  gsea_df <- as.data.frame(gsea_results[[s]]$GO)

  # STEP 2: Sort by NES
  gsea_df <- gsea_df %>%
    arrange(desc(NES))

  # STEP 3: Select and format key columns
  filtered_df <- gsea_df %>%
    select(ID, Term = Description, adj_pval = p.adjust, core_enrichment) %>%
    mutate(Category = "BP") %>%
    relocate(Category)

  # STEP 4: Map ENTREZ IDs to SYMBOLs
  filtered_df$Genes <- str_split(filtered_df$core_enrichment, "/")
  all_entrez <- unique(unlist(filtered_df$Genes))
  entrez2symbol <- bitr(all_entrez, fromType = "ENTREZID", toType = "SYMBOL", OrgDb = org.Hs.eg.db)
  entrez_map <- setNames(entrez2symbol$SYMBOL, entrez2symbol$ENTREZID)

  filtered_df$Genes <- lapply(filtered_df$Genes, function(g) {
    gene_symbols <- entrez_map[g]
    gene_symbols <- gene_symbols[!is.na(gene_symbols)]
    paste(gene_symbols, collapse = ", ")
  })

  final_df <- filtered_df %>%
    select(Category, ID, Term, Genes, adj_pval)
  
   #  STEP 7: Extract DE data
de_df <- df_list[[s]] %>%
  filter(!is.na(symbol)) %>%
  select(ID = symbol, logFC = log2FoldChange) %>%
  distinct(ID, .keep_all = TRUE)

  

  # STEP 5: Circle data
  circ <- circle_dat(final_df, de_df)

  # STEP 6: Top terms and genes
  top_terms <- final_df %>%
    slice(1:5) %>%
    pull(Term)

  top_genes <- final_df %>%
    slice(1:5) %>%
    pull(Genes) %>%
    str_split(",\\s*") %>%
    unlist() %>%
    unique()

 filtered_de_df <- de_df %>%
    filter(ID %in% top_genes)

  # STEP 8: Chord plot
  chord <- chord_dat(circ, filtered_de_df, top_terms)

  # STEP 9: Plot with title
  print(
  GOChord(chord,
          limit = c(3, 0),
          gene.order = "logFC",
          title = paste("GOChord Plot -", s))
)
}

# Close PDF
dev.off()
```

# visualizing gsea results as whole
```{r}
# === 1. Prepare full GSEA result summary ===
gsea_summary_all <- do.call(rbind, lapply(names(gsea_results), function(name) {
  res_list <- list()
  
  for (source in c("GO", "KEGG")) {
    gsea <- gsea_results[[name]][[source]]
    
    if (!is.null(gsea) && nrow(as.data.frame(gsea)) > 0) {
      df <- as.data.frame(gsea)
      df$Condition <- name
      df$Source <- source
      df <- df[, c("Condition", "Source", "ID", "Description", "p.adjust", "NES")]
      res_list[[source]] <- df
    }
  }
  
  do.call(rbind, res_list)
})) %>%
  mutate(
    NES = as.numeric(NES),
    negLogPadj = -log10(p.adjust)
  )

# === 2. Top 5 Positively Enriched ===
top_pos <- gsea_summary_all %>%
  filter(NES > 0) %>%
  filter(p.adjust<0.05) %>%
  group_by(Condition, Source) %>%
  slice_max(order_by = NES, n = 5, with_ties = FALSE) %>%
  ungroup()

# === 3. Top 5 Negatively Enriched ===
top_neg <- gsea_summary_all %>%
  filter(NES < 0) %>%
   filter(p.adjust<0.05) %>%
  group_by(Condition, Source) %>%
  slice_min(order_by = NES, n = 5, with_ties = FALSE) %>%
  ungroup()

# === 4. Calculate global color scale limits ===
min_logpadj <- min(c(top_pos$negLogPadj, top_neg$negLogPadj), na.rm = TRUE)
max_logpadj <- max(c(top_pos$negLogPadj, top_neg$negLogPadj), na.rm = TRUE)

# === 5. Updated plot function with locked color scale ===
bubble_plot <- function(data, title) {
  ggplot(data, aes(x = Condition, y = reorder(Description, negLogPadj))) +
    geom_point(aes(size = abs(NES), color = negLogPadj)) +
    scale_color_gradient(low = "blue", high = "red", limits = c(min_logpadj, max_logpadj)) +
    labs(title = title,
         x = "Condition", y = "Pathway",
         size = "|NES|", color = "-log10(p.adjust)") +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text.y = element_text(size = 9)
    )
}

# === 5. Individual Plots ===
plot_go_pos <- bubble_plot(top_pos %>% filter(Source == "GO"), "Top Positively Enriched GO Terms")
plot_kegg_pos <- bubble_plot(top_pos %>% filter(Source == "KEGG"), "Top Positively Enriched KEGG Pathways")
plot_go_neg <- bubble_plot(top_neg %>% filter(Source == "GO"), "Top Negatively Enriched GO Terms")
plot_kegg_neg <- bubble_plot(top_neg %>% filter(Source == "KEGG"), "Top Negatively Enriched KEGG Pathways")

# === 6. Arrange 2x2 Grid ===
(plot_go_pos | plot_kegg_pos) / 
(plot_go_neg | plot_kegg_neg)


```