---
title: "H2171_GSEA_TEAD"
author: "Nivitha Bhaskar"
date: "2025-05-29"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Loading_libraries}
library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
library(AnnotationDbi)
library(ggplot2)
library(enrichplot)
library(GO.db)
library(GOplot)
library(enrichplot)
library(tidyverse)
library(ggvenn)
library(pheatmap)
```

# GSEA with custom TEAD gene set
```{r}
# Load and clean the custom new gene sets
tead_new <- read.csv("common_TEAD_genes_H2171.csv")

custom_gene_set_new <- list("TEAD_repressed" = tead_new$TEAD_repressed)
custom_induced_new <- list("TEAD_induced" = tead_new$TEAD_induced)

# Clean out any empty strings
custom_gene_set_new <- lapply(custom_gene_set_new, function(x) x[x != ""])
custom_induced_new <- lapply(custom_induced_new, function(x) x[x != ""])

# Convert to TERM2GENE format
term2gene_repressed_new <- stack(custom_gene_set_new)
colnames(term2gene_repressed_new) <- c("gene", "term")
term2gene_repressed_new <- term2gene_repressed_new[, c("term", "gene")]

term2gene_induced_new <- stack(custom_induced_new)
colnames(term2gene_induced_new) <- c("gene", "term")
term2gene_induced_new <- term2gene_induced_new[, c("term", "gene")]

```

# Loading differential analysis results
```{r}
# List of the file names
file_names <- c("H2171_corin_ic50_final.csv", "H2171_corin_ic75_final.csv", 
                "H2171_ent_ic50_final.csv", "H2171_ent_ic75_final.csv", 
                "H2171_sp_ic50_final.csv", "H2171_sp_ic75_final.csv")

# Read all the CSV files into a list of data frames
df_list <- lapply(file_names, read.csv)

# Optionally, assign names to each element in the list for easy reference
names(df_list) <- c("H2171_corin_ic50", "H2171_corin_ic75", "H2171_ent_ic50", "H2171_ent_ic75", "H2171_sp_ic50", "H2171_sp_ic75")

# You can access each data frame by its name in the list eg.
df_list$H2171_corin_ic50
```

# custom GSEA function to loop through all the datasets
```{r}
run_custom_gsea_new <- function(df) {
  # Use gene symbols directly from the last column
  gene_symbols <- as.character(df[[ncol(df)]])
  gene_list <- df[[3]]  # assuming logFC or stat is in column 3
  names(gene_list) <- gene_symbols

  # Clean gene list
  gene_list <- sort(gene_list, decreasing = TRUE)
  gene_list <- gene_list[!duplicated(names(gene_list))]
  gene_list <- gene_list[!is.na(names(gene_list))]

  # Run GSEA for both repressed and induced gene sets
  gsea_repressed <- tryCatch({
    GSEA(geneList = gene_list, TERM2GENE = term2gene_repressed_new, pvalueCutoff = 0.05)
  }, error = function(e) NULL)

  gsea_induced <- tryCatch({
    GSEA(geneList = gene_list, TERM2GENE = term2gene_induced_new, pvalueCutoff = 0.05)
  }, error = function(e) NULL)

  return(list(repressed = gsea_repressed, induced = gsea_induced))
}

gsea_custom_results_new <- lapply(df_list, run_custom_gsea_new)
names(gsea_custom_results_new) <- names(df_list)
```

```{r}
pdf("H2171_GSEA_TEAD.pdf", width = 20, height = 20)

# Loop through each dataset
for (name in names(gsea_custom_results_new)) {
  res_list <- gsea_custom_results_new[[name]]

  # Plot TEAD_repressed
  if (!is.null(res_list$repressed) && nrow(res_list$repressed@result) > 0) {
    plot_r <- gseaplot(res_list$repressed, geneSetID = 1, title = paste(name, "TEAD_repressed"))
    print(plot_r)  # Ensure it goes into the PDF
  }

  # Plot TEAD_induced
  if (!is.null(res_list$induced) && nrow(res_list$induced@result) > 0) {
    plot_i <- gseaplot(res_list$induced, geneSetID = 1, title = paste(name, "TEAD_induced"))
    print(plot_i)  # Ensure it goes into the PDF
  }
}

dev.off()

```

# summarising the result

```{r}
# Create an empty list to store results
summary_list_new <- list()

# Loop through each dataset and extract info
for (name in names(gsea_custom_results_new)) {
  res_list <- gsea_custom_results_new[[name]]
  
  # Repressed results
  if (!is.null(res_list$repressed) && nrow(res_list$repressed@result) > 0) {
    top_repressed <- res_list$repressed@result[1, c("ID", "NES", "p.adjust")]
    top_repressed$Dataset <- name
    top_repressed$Type <- "TEAD_repressed"
    summary_list_new[[paste0(name, "_repressed")]] <- top_repressed
  }
  
  # Induced results
  if (!is.null(res_list$induced) && nrow(res_list$induced@result) > 0) {
    top_induced <- res_list$induced@result[1, c("ID", "NES", "p.adjust")]
    top_induced$Dataset <- name
    top_induced$Type <- "TEAD_induced"
    summary_list_new[[paste0(name, "_induced")]] <- top_induced
  }
}

# Combine into one data frame
summary_df_new <- do.call(rbind, summary_list_new)

# Reorder columns for clarity
summary_df_new <- summary_df_new[, c("Dataset", "Type", "ID", "NES", "p.adjust")]

# View the table
print(summary_df_new)
```

# extracting the leading edge genes contributing to custom GSEA

```{r}
# Create a list to store leading edge genes
leading_edge_genes_list <- list()

for (name in names(gsea_custom_results_new)) {
  res_list <- gsea_custom_results_new[[name]]
  
  # Repressed
  if (!is.null(res_list$repressed) && nrow(res_list$repressed@result) > 0) {
    genes <- strsplit(res_list$repressed@result$core_enrichment, "/")[[1]]
    leading_edge_genes_list[[paste0(name, "_repressed")]] <- genes
  }

  # Induced
  if (!is.null(res_list$induced) && nrow(res_list$induced@result) > 0) {
    genes <- strsplit(res_list$induced@result$core_enrichment, "/")[[1]]
    leading_edge_genes_list[[paste0(name, "_induced")]] <- genes
  }
}

# Check one of them
leading_edge_genes_list$H2171_corin_ic50_repressed
```

# Over-representational analysis in R - not trimmed
```{r}
# Convert gene symbols to Entrez IDs
convert_to_entrez <- function(genes) {
  bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)$ENTREZID
}

# Store ORA results
ora_results_list <- list()

# Loop through each leading edge gene set
for (name in names(leading_edge_genes_list)) {
  genes <- leading_edge_genes_list[[name]]
  entrez_genes <- convert_to_entrez(genes)

  if (length(entrez_genes) > 5) {  # Need a few genes for enrichment
    # KEGG pathway enrichment
    enrich_res <- enrichKEGG(gene = entrez_genes,
                             organism = 'hsa',
                             pAdjustMethod = "BH",
                             pvalueCutoff = 0.05)
    ora_results_list[[name]] <- enrich_res
  }
}
```
# Viewing over-represented pathways - not trimmed
```{r}

# Combine enrichResult objects into a clean summary table
ora_summary <- do.call(rbind, lapply(names(ora_results_list), function(name) {
  enrich_res <- ora_results_list[[name]]
  
  if (!is.null(enrich_res) && nrow(as.data.frame(enrich_res)) > 0) {
    df <- as.data.frame(enrich_res)
    df$Condition <- name  # Add condition label
    # Some enrichResult objects may not have zScore; add NA if missing
    if (!"zScore" %in% colnames(df)) df$zScore <- NA
    return(df[, c("Condition", "ID", "Description", "p.adjust", "zScore")])
  }
  return(NULL)
}))

# View the table
head(ora_summary)

# Ensure zScore and p.adjust are numeric
ora_summary$zScore <- as.numeric(ora_summary$zScore)
ora_summary$negLogPadj <- -log10(ora_summary$p.adjust)

# Bubble plot
ggplot(ora_summary, aes(x = Condition, y = Description)) +
  geom_point(aes(size = zScore, color = negLogPadj)) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 10)) +
  labs(title = "Enriched Pathways by Condition",
       x = "Condition",
       y = "Pathway",
       color = "-log10(p.adjust)",
       size = "zScore")
```

# Running the custom gsea with trimmed top 3000 and bottom 3000 genes
```{r}
run_custom_gsea_trimmed <- function(df) {
  # Step 1: Extract gene symbols and ranking metric
  gene_symbols <- as.character(df[[ncol(df)]])
  gene_stat <- sign(df[[3]]) * -log10(df[[7]])  # assuming logFC or stat is in column 3
  
  # Step 2: Combine and sort
  gene_df <- data.frame(symbol = gene_symbols, stat = gene_stat)
  gene_df <- gene_df[!duplicated(gene_df$symbol) & !is.na(gene_df$symbol), ]
  gene_df <- gene_df[order(gene_df$stat, decreasing = TRUE), ]
  
  # Step 3: Keep top 200 and bottom 200 genes
  trimmed_df <- rbind(
    head(gene_df, 2000),
    tail(gene_df, 2000)
  )
  
  
  # Step 4: Create named vector for GSEA
  gene_list_trimmed <- trimmed_df$stat
  names(gene_list_trimmed) <- trimmed_df$symbol
  
  # Step 5: Run GSEA for both repressed and induced gene sets
  gsea_repressed <- tryCatch({
    GSEA(geneList = sort(gene_list_trimmed, decreasing = TRUE), 
         TERM2GENE = term2gene_repressed_new, pvalueCutoff = 0.05)
  }, error = function(e) NULL)
  
  gsea_induced <- tryCatch({
    GSEA(geneList = sort(gene_list_trimmed, decreasing = TRUE), 
         TERM2GENE = term2gene_induced_new, pvalueCutoff = 0.05)
  }, error = function(e) NULL)
  
  return(list(repressed = gsea_repressed, induced = gsea_induced))
}

gsea_trimmed_results_H2171 <- lapply(df_list, run_custom_gsea_trimmed)
names(gsea_trimmed_results_H2171) <- names(df_list)
```

```{r}
pdf("H2171_GSEA_TEADtrimmed.pdf", width = 20, height = 20)

# Loop through each dataset
for (name in names(gsea_trimmed_results_H2171)) {
  res_list <- gsea_trimmed_results_H2171[[name]]

  # Save TEAD_repressed plot
  if (!is.null(res_list$repressed) && nrow(res_list$repressed@result) > 0) {
    plot_r <- gseaplot(res_list$repressed, geneSetID = 1, title = paste(name, "TEAD_repressed"))
    print(plot_r)
  }

  # Save TEAD_induced plot
  if (!is.null(res_list$induced) && nrow(res_list$induced@result) > 0) {
    plot_i <- gseaplot(res_list$induced, geneSetID = 1, title = paste(name, "TEAD_induced"))
    print(plot_i)
  }
}
```

# Summary of trimmed result
```{r}
# Create an empty list to store results
summary_trimmed_H2171 <- list()

# Loop through each dataset and extract info
for (name in names(gsea_trimmed_results_H2171)) {
  res_list <- gsea_trimmed_results_H2171[[name]]
  
  # Repressed results
  if (!is.null(res_list$repressed) && nrow(res_list$repressed@result) > 0) {
    top_repressed <- res_list$repressed@result[1, c("ID", "NES", "p.adjust")]
    top_repressed$Dataset <- name
    top_repressed$Type <- "TEAD_repressed"
    summary_trimmed_H2171[[paste0(name, "_repressed")]] <- top_repressed
  }
  
  # Induced results
  if (!is.null(res_list$induced) && nrow(res_list$induced@result) > 0) {
    top_induced <- res_list$induced@result[1, c("ID", "NES", "p.adjust")]
    top_induced$Dataset <- name
    top_induced$Type <- "TEAD_induced"
    summary_trimmed_H2171[[paste0(name, "_induced")]] <- top_induced
  }
}

# Combine into one data frame
summary_dftrim_H2171 <- do.call(rbind, summary_trimmed_H2171)

# Reorder columns for clarity
summary_dftrim_H2171 <- summary_dftrim_H2171[, c("Dataset", "Type", "ID", "NES", "p.adjust")]

# View the table
print(summary_dftrim_H2171)
```

# Barplot of the NES scores
```{r}
# Your data (assuming already loaded into df_trim)
# Add significance stars
df_trim <- summary_dftrim_H2171 %>%
  mutate(Significance = case_when(
    p.adjust < 0.001 ~ "***",
    p.adjust < 0.01 ~ "**",
    p.adjust < 0.05 ~ "*",
    TRUE ~ ""
  ))

# Create the bar plot
ggplot(df_trim, aes(x = reorder(Dataset, NES), y = NES, fill = Type)) +
  geom_col(width = 0.7, position = position_dodge()) +
  geom_text(aes(label = Significance),
            vjust = ifelse(df_trim$NES > 0, -0.5, 1.2),
            size = 4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  coord_flip() +
  scale_fill_manual(values = c("TEAD_repressed" = "#1b9e77", "TEAD_induced" = "#d95f02")) +
  labs(title = "GSEA Normalized Enrichment Scores (NES)",
       x = "Condition",
       y = "NES",
       fill = "Gene Set") +
  theme_light() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.y = element_text(size = 10)
  )
```

# Analysing the leading genes contributing to this enrichment

```{r}
# Create a list to store leading edge genes
leading_edge_genes_list_t <- list()

for (name in names(gsea_trimmed_results_H2171)) {
  res_list <- gsea_trimmed_results_H2171[[name]]
  
  # Repressed
  if (!is.null(res_list$repressed) && nrow(res_list$repressed@result) > 0) {
    genes <- strsplit(res_list$repressed@result$core_enrichment, "/")[[1]]
    leading_edge_genes_list_t[[paste0(name, "_repressed")]] <- genes
  }

  # Induced
  if (!is.null(res_list$induced) && nrow(res_list$induced@result) > 0) {
    genes <- strsplit(res_list$induced@result$core_enrichment, "/")[[1]]
    leading_edge_genes_list_t[[paste0(name, "_induced")]] <- genes
  }
}

# Check one of them
leading_edge_genes_list_t$H2171_corin_ic50_repressed
```

# Pathway over-representation analysis (ORA) with leading edge genes

```{r}
# Convert gene symbols to Entrez IDs
convert_to_entrez <- function(genes) {
  bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)$ENTREZID
}

# Store ORA results
ora_results_list_t <- list()

# Loop through each leading edge gene set
for (name in names(leading_edge_genes_list_t)) {
  genes <- leading_edge_genes_list_t[[name]]
  entrez_genes <- convert_to_entrez(genes)

  if (length(entrez_genes) > 5) {  # Need a few genes for enrichment
    # KEGG pathway enrichment
    enrich_res <- enrichKEGG(gene = entrez_genes,
                             organism = 'hsa',
                             pAdjustMethod = "BH",
                             pvalueCutoff = 0.05)
    ora_results_list_t[[name]] <- enrich_res
  }
}
```

# Viewing over-represented pathways
```{r}

# Combine enrichResult objects into a clean summary table
ora_summary <- do.call(rbind, lapply(names(ora_results_list_t), function(name) {
  enrich_res <- ora_results_list_t[[name]]
  
  if (!is.null(enrich_res) && nrow(as.data.frame(enrich_res)) > 0) {
    df <- as.data.frame(enrich_res)
    df$Condition <- name  # Add condition label
    # Some enrichResult objects may not have zScore; add NA if missing
    if (!"zScore" %in% colnames(df)) df$zScore <- NA
    return(df[, c("Condition", "ID", "Description", "p.adjust", "zScore")])
  }
  return(NULL)
}))

# View the table
head(ora_summary)

# Ensure zScore and p.adjust are numeric
ora_summary$zScore <- as.numeric(ora_summary$zScore)
ora_summary$negLogPadj <- -log10(ora_summary$p.adjust)


ggplot(ora_summary, aes(x = Condition, y = Description)) +
  geom_point(aes(size = zScore, color = negLogPadj)) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_light() +
  theme(axis.text.y = element_text(size = 10)) +
  labs(title = "Enriched Pathways by Condition",
       x = "Condition",
       y = "Pathway",
       color = "-log10(p.adjust)",
       size = "zScore")

```